### 2004-07-16 doc/Simulation_and_Control_using_Supremica/automationobjects.tex
*** empty log message ***

%% Background
With function blocks the main technological problem of
distributed automation has been solved. A general framework
for development and deployment of distributed automation
system has been developed and established. What is still
needed is a way of encapsulating function blocks together
with the data about their properties and, more importantly,
their functionality in a way that will enable diminishing of
the time to market and facilitate effective rapid object
oriented development of automation systems.
%% Why automation objects? What do we want to do?
After considering the following arguments it will become
clear why such encapsulation is needed and how it can help
in achieving the afore mentioned goal.

%%     effective reuse of software
Object oriented system development as presented in
\cite{iec:614991:2000} is not enough for achieving
\textit{effective reuse of software} in automation systems
development. This is the case because effective software
reuse demands efficient exchange of information about how
the software that is supposed to be reused is used and what
it does. In other words, interfaces of the software objects
have to be well specified and well documented. The
\cite{iec:614991:2000} makes sure that the interfaces are
well specified but it does not provide anything that will
assure that they are well documented.

@@

%% Definition. 
%%     multi view objects
%%     view is the interface exposed for different purposes.
An \textbf{automation object} is defined as a set of
applications, subapplications and function blocks (as
defined in \cite{iec:614991:2000}) encapsulated with a
declaration of \textit{attributes} and \textit{views} of
these.

An \textbf{attribute} is user defined and specified data that
describes some significant property of the automation
object.

A \textbf{view} is a description of a certain aspect of the
automation object and information about how that aspect is
used in the automation system.
% view is functional description and its name should only
% contain what is needed to understand its function
% eg elevator, pump, oven, sensor
% all the information about what hardware this automation
% block is for should be in the attributes
% eg attribute hardware='abb robot N25'

@@

% Support in Supremica

>>>>

%% Intro
With the function blocks the main technological problem of
distributed automation has been solved. A general framework
for development and deployment of distributed automation
system has been developed and established. What is still
needed is a way of encapsulating function blocks together
with the data about their properties and, more importantly,
their functionality in a way that will enable diminishing of
the time to market and facilitate effective rapid object
oriented development of automation systems.

\section{Background}
%% Background. Why automation objects? What do we want to do?
Considering the following arguments it is clear why such
encapsulation is needed and how it can help in achieving the
afore mentioned goal.

%%     effective reuse of software
Object oriented system development as presented in the
IEC~61499 standard is ideal for the purpose that is is
designed for but not enough for achieving {\it effective
  software reuse} in automation systems development.

% grouping software objects in functional units (encapsulation)
% abstraction of functionality (views)
% properties of the functional unit (attributes)

@@

\section{Definition}
%% Definition. 
%%     multi view objects
An \textbf{automation object} is defined as \textit{a set of
  software objects} encapsulated with a declaration of
\textit{services}, \textit{attributes} and \textit{views} of
the set.

The \textbf{set} specifies the applications, subapplications
and function blocks, as defined in \cite{iec:614991:2000},
contained within the automation object. Only one set per
automation object is defined and it contains \textit{all} of
the software objects required for the functionality of the
automation object.
% less dependency problems at the cost of storage, more on
% this after the definition

An \textbf{attribute} is a user defined and specified data that
describes some significant property of the automation
object.

A \textbf{view} is an interface to the object. Every
interface has at least one implementation.

A \textbf{service}

@@

\section{Implementation and Support in Supremica}
% Support in Supremica

### 2004-07-16 doc/Simulation_and_Control_using_Supremica/functionblocks.tex
*** empty log message ***

\chapter{IEC 61499 Function Block Standard}
\label{functionblocks}

\begin{bf}
{This chapter is with minor revisions copied from chapter 2 in \cite{j:des:2002}.}
\end{bf}

New technologies and standards are emerging that are set to have a
dramatic effect on the design and implementation of industrial
control systems. A new standard IEC-61131-3~\cite{iec:1131:1993} was
introduced in 1995 to standardize the representation of control
designs so as to enable cross-vendor integration. The standard
offered significant advantages over earlier vendor proprietary
control design representations~\cite{l:pro:1995}.
\begin{itemize}
\item It encouraged well-structured procedural
programs that allowed the development of different functional
elements into separate procedures or program units.
\item It provided standardized languages and methods of program execution
so that wide range of technological problems can be programmed as
vendor-independent software elements.
\item It provided support for various languages like structured
text, function block diagrams, ladder diagrams, instruction list
and sequential function charts.
\item The standard enforced strong data typing and supported user defined
data structures.
\end{itemize}

@@

To overcome these problems, a function block model was sought that
would support distributed control systems in a scalable,
extensible and flexible manner. IEC-61499~\cite{iec:61499:2000}
standard has now been developed for distributed industrial
process, measurement and control systems.

@@

The following sections are adapted largely from the various
discussions on the features of the standard in ~\cite{iec:61499:2000,
iec:1131:1993, l:pro:1995, l:mod:2001, c:des:2002, c:ope:2002}

\section{Reference Models}
IEC-61499 has provided a set of models for describing distributed
systems that have been programmed using function blocks.
\begin{itemize}
\item {\bf System Model:} At the physical level, a distributed
system consists of a set of co-operating {\it applications}. An
application can exist on a single system such as application C in
figure \ref{f:System_Model} or have functionality distributed over
a number of devices such as applications A and B in Figure
\ref{f:System_Model}.

\item {\bf Device Model:} A resource provides independent
execution and control of networks of function blocks. It supports
zero or more {\it resources} and has at least one interface. The
interface could either be a 'process interface' that allows the
resources to exchange data with the input and output points on the
physical devices, or a communication interface that allows the
resources to communicate with external resources in remote
devices. The device model is illustrated in Figure
\ref{f:Device_Model}.

\item {\bf Resource Model:} A resource is an independent functional unit,
contained in a device which has independent control of its
operation. The resource can be created, configured, parameterized,
started up, deleted etc. without effecting other resources. Each
resource provides an interface to the communication systems and
tot the device specific process. The resource is therefore,
concerned with the mapping of data and event flows which pass
between function blocks in the local resource to remove resource
function blocks via the device communication interfaces.

@@

\item {\bf Application Model:} An application is defined as a
network of interconnected function blocks, linked by data and
event flows. An application can be fragmented and distributed over
many resources. The resources ensure that the events are executed
in the correct serial order based on the causal relationships
specified by the applications.

@@

\item {\bf Function Block Model:} A function block is a functional
unit of software that its own data structure which can be
manipulated by one or more algorithms. The formal description of
the data structure is provided by the function block type
definition.

@@

\end{itemize}

%
%%\singlespace
\begin{figure}
\begin{center}
\epsfig{file=images/functionblocks/System_Model.eps, scale=.8} \caption[System
Model]{System Model {\protect ~\cite{iec:61499:2000}}}
\label{f:System_Model}
\end{center}
\end{figure}
%%\doublespace
%

%
%%\singlespace
\begin{figure}
\begin{center}
\epsfig{file=images/functionblocks/Device_Model.eps, scale=.8} \caption[Device
Model]{Device Model{\protect ~\cite{iec:61499:2000}}}
\label{f:Device_Model}
\end{center}
\end{figure}
%%\doublespace
%

%
%%\singlespace
\begin{figure}
\begin{center}
\epsfig{file=images/functionblocks/Resource_Model.eps, scale=.8} \caption[Resource
Model]{Resource Model{\protect ~\cite{iec:61499:2000}}}
\label{f:Resource_Model}
\end{center}
\end{figure}
%%\doublespace
%

%
%%\singlespace
\begin{figure}
\begin{center}
\epsfig{file=images/functionblocks/Application_Model.eps, scale=.8}
\caption[Application Model]{Application Model{\protect
~\cite{iec:61499:2000}}} \label{f:Application_Model}
\end{center}
\end{figure}
%%\doublespace
%

%
%%\singlespace
\begin{figure}
\begin{center}
\epsfig{file=images/functionblocks/Function_Block_Char.eps, scale=.8}
\caption[Function Block Characteristics]{Function Block
Characteristics{\protect ~\cite{iec:61499:2000}}}
\label{f:Function_Block_Char}
\end{center}
\end{figure}
%%\doublespace
%

\section{Function Block Types}

A function block type is defined by a type name, formal
definitions for the input events and data and the output events
and data. The type definition also includes the internal behavior
of the block. The notion of a function block type is analogous to
the class definition in object-oriented programming while function
block instances correspond to the object instances of the classes.

\begin{itemize}
\item{\bf Basic Function Block Types:} The behavior of basic
function blocks is defined in terms of algorithms that are invoked
in response to input events. As algorithms execute, they trigger
output events to signal their state to other blocks. The basic
blocks hold their state in terms of internal variables. Basic
function blocks also can communicate their state using input and
output data variables.

@@

\item{\bf Composite Function Blocks:} The behavior of a composite
function block is described in terms of the event and data
connections between its component function blocks. A composite
function block can be used to hide the internal interactions
between the component blocks and presenting a simplified interface
for use by the outside world. A composite function block is
illustrated in Figure \ref{f:Composite_Block}.

\item {\bf Service Interface Function Blocks:} Service interface
function blocks provide an interface between the function block
domain and the external services. Since a service interface
function block is primarily concerned with data exchange, it is
defined using time sequence diagrams. Examples of service
interface function blocks are communication blocks for publishing
and subscribing over a network and management blocks for
dynamically creating component blocks and event connections.

\item {\bf Event Function Blocks:} Event function blocks are some
special basic function blocks defined in the standard to ease the
management of events. These blocks are used for control,
generation and detection of events and can be used by other
composite blocks. These blocks are defined using an execution
control chart or a time sequence diagram. These blocks can be
used, for instance, for the following purposes.
\begin{enumerate}
\item To split events to produce new events
\item To allow events to be merged.
\item To permit the propagation of certain events.
\item To select between two or more events.
\item To delay an event by a given period.
\item To generate a stream of events at a regular interval.
\end{enumerate}

\end{itemize}

%
%\singlespace
\begin{figure}
\begin{center}
\epsfig{file=images/functionblocks/Basic_Function_Block.eps, scale=.8}
\caption[Basic Function Block]{Basic Function Block{\protect
~\cite{iec:61499:2000}}} \label{f:Basic_Block}
\end{center}
\end{figure}
%\doublespace
%

%
%\singlespace
\begin{figure}
\begin{center}
\epsfig{file=images/functionblocks/Composite_Function_Block.eps, scale=.8}
\caption[Composite Function Block]{Composite Function
Block{\protect ~\cite{iec:61499:2000}}} \label{f:Composite_Block}
\end{center}
\end{figure}
%\doublespace
%

\section{Execution Model}

The execution of algorithms for basic function blocks is invoked
by the execution control portion of a function block instance in
response to events at event inputs. This invocation takes the form
of a request to the underlying scheduling function of the
associated resource to schedule the execution of algorithm's
operations. The model assumes that the resource in which a
function block exists provides a scheduling function that ensures
each phase of function block execution occurs in the correct order
and with the correct priority.

@@

%
%\singlespace
\begin{figure}
\begin{center}
\epsfig{file=images/functionblocks/Execution_Control_Chart.eps, scale=.8}
\caption[Execution Model]{Execution Model{\protect
~\cite{iec:61499:2000}}} \label{f:Execution_Model}
\end{center}
\end{figure}
%\doublespace
%

@@

\section{FBEditor}
Function Block Development Kit (FBDK) ~\cite{c:fun:2002} is a
software toolkit that enables users to build and test data types,
function block types, resource types, device types and system
configurations according to the IEC 61499 standard. FBEditor is a
part of the FBDK toolkit that provides a development and runtime
environment for applications based on the IEC 61499 standard.
Figure \ref{f:FBEditor} shows the editor being used to load a
function block used for controlling a motor.

%
%\singlespace
\begin{figure}
\begin{center}
\epsfig{file=images/functionblocks/FBEditor.eps, scale=.6}
\caption[FBEditor]{FBEditor{\protect ~\cite{c:fun:2002}}}
\label{f:FBEditor}
\end{center}
\end{figure}
%\doublespace
%

FBEditor enables the users to edit the block definitions through
both a text editor and through a graphical user interface. New
data types, basic function blocks, composite function blocks,
execution control charts and other higher level entities may be
created through the editor. The attributes of the blocks (data and
events) may also be examined through a graphical user interface.
The function blocks can be examined at the device level as shown
in Figure \ref{f:Editor-Devices} or at the level of individual
attributes like the connections, as shown in figure
\ref{f:Editor-Mapping}.

The editor supports the languages supported by the standard
(Structured text, execution control chart etc.). It uses XML as
the exchange format to ensure portability across various domains.
Figure \ref{f:Editor-Versions} shows how the editor can also be
used to manage meta properties like versions, owner entities,
authors etc.

%
%\singlespace
\begin{figure}
\begin{center}
\epsfig{file=images/functionblocks/Editor-Devices.eps, scale=.6} \caption[FBEditor
- Editing system devices]{FBEditor - Editing system
devices{\protect ~\cite{c:fun:2002}}} \label{f:Editor-Devices}
\end{center}
\end{figure}
%\doublespace
%

%
%\singlespace
\begin{figure}
\begin{center}
\epsfig{file=images/functionblocks/Editor-Mapping.eps, scale=.6} \caption[FBEditor
- Editing device connections]{FBEditor - Editing device
connections{\protect ~\cite{c:fun:2002}}} \label{f:Editor-Mapping}
\end{center}
\end{figure}
%\doublespace
%

%
%\singlespace
\begin{figure}
\begin{center}
\epsfig{file=images/functionblocks/Editor-Versions.eps, scale=.6}
\caption[FBEditor - Editing meta properties]{FBEditor - Editing
meta properties{\protect ~\cite{c:fun:2002}}}
\label{f:Editor-Versions}
\end{center}
\end{figure}
%\doublespace
%
FBEditor has a navigation pane that allows the user to choose the
display on the worksheet area, including the external interface;
the Execution Control Chart (ECC) for basic function block type;
service primitive sequences for service interface function block
types; and function block networks for composite function block
types, device and resource types and instances, and system
configurations. The worksheet / viewport area contains the
function block diagram for the currently selected instance.
Different components of the diagram can be dragged around for
better readability.

FBEditor has been developed in Java language. Hence in order to
run simulations on applications written using the editor, one
needs to generate the Java classes that correspond to the source
files for the function blocks written in structured text (for
instance). The editor has a code generation mechanism through
which it parses through the source files and generates a Java
class corresponding to each block.

On occurrence of certain input events, the user needs to specify
the algorithms that should be scheduled to be executed and the
output events that should signal the completion of algorithms.
This may be done either by editing the generated Java files or by
enclosing the algorithm inside especially marked tags in the
source files. The files are then compiled and simulations can be
run from within the editor environment. Figure
\ref{f:Editor-Transitions} shows the assignment of output events
on the transitions through the editing of the execution control
chart.

%
%\singlespace
\begin{figure}
\begin{center}
\epsfig{file=images/functionblocks/Editor-Transitions.eps, scale=.6}
\caption[FBEditor Editing the execution control chart]{FBEditor
Editing the execution control chart{\protect ~\cite{c:fun:2002}}}
\label{f:Editor-Transitions}
\end{center}
\end{figure}
%\doublespace
%

%\section{EMBench and Function Blocks}

%Since the FBEditor provided the aforementioned functionality, it
%was decided that the EMBench should leverage it. In particular,
%the bench relies on the editor for two basic features.
%\begin{enumerate}
%\item To generate the Java files from the IEC 61499 source files.
%\item To manage the Java thread that initializes the blocks during
%simulation.
%\end{enumerate}

%It has been visualized that the bench will eventually have the
%editor as one of its components. The bench explicitly manages the
%design activities of the user and generates the function block
%definitions for the user design. It then requests the FBEditor
%component to load the specified source files, generate the Java
%classes and run the simulation. In addition, the user may make
%behavioral changes to the blocks by editing the events, if he/she
%so desires. The editor conforms to the published service
%specifications of the bench as explained in chapter \ref{chap3}
%and hence would present the same interface to the client.

%Currently, however, the FBEditor is not a part of the bench
%because it does not expose an API. Since it is external to the
%bench, the bench has to circumvent this problem by allowing the
%editor to be executed as a separate application and delegating the
%simulation task completely to the editor. The data exchange
%necessary between the two applications is handled via network
%communication. This issue is explained in a greater detail in
%chapter~\ref{chap4} and chapter~\ref{chap5}.

>>>>

\chapter{IEC~61499 Standard}
\label{functionblocks}

\textbf{This chapter,with minor revisions, is copied from
  chapter 2 in \cite{j:des:2002}.}

New technologies and standards are emerging that are set to
have a dramatic effect on the design and implementation of
industrial control systems. A new standard
IEC~61131-3~\cite{iec:1131:1993} was introduced in 1995 to
standardize the representation of control designs so as to
enable cross-vendor integration. The standard offered
significant advantages over earlier vendor proprietary
control design representations~\cite{l:pro:1995}.
\begin{itemize}
\item It encouraged well-structured procedural programs that
  allowed the development of different functional elements
  into separate procedures or program units.
\item It provided standardized languages and methods of
  program execution so that wide range of technological
  problems can be programmed as vendor-independent software
  elements.
\item It provided support for various languages like
  structured text, function block diagrams, ladder diagrams,
  instruction list and sequential function charts.
\item The standard enforced strong data typing and supported
  user defined data structures.
\end{itemize}

@@

To overcome these problems, a function block model was
sought that would support distributed control systems in a
scalable, extensible and flexible manner.
IEC~61499-1~\cite{iec:614991:2000} standard has now been
developed for distributed industrial process, measurement
and control systems.

@@

The following sections are adapted largely from the various
discussions on the features of the standard in
~\cite{iec:614991:2000, iec:1131:1993, l:pro:1995,
  l:mod:2001, c:des:2002, c:ope:2002}

\section{Reference Models}
IEC~61499-1 has provided a set of models for describing
distributed systems that have been programmed using function
blocks.

\subsection{System Model} 
At the physical level, a distributed system consists of a
set of co-operating {\it applications}. An application can
exist on a single system such as application C in figure
\ref{f:System_Model} or have functionality distributed over
a number of devices such as applications A and B in Figure
\ref{f:System_Model}.

\begin{figure}
  \begin{center}
    \epsfig{file=images/functionblocks/System_Model.eps,scale=.8} 
    \caption[System Model]
            {System Model{\protect ~\cite{iec:614991:2000}}}
    \label{f:System_Model}
  \end{center}
\end{figure}

\subsection{Device Model} 
A resource provides independent execution and control of
networks of function blocks. It supports zero or more {\it
  resources} and has at least one interface. The interface
could either be a 'process interface' that allows the
resources to exchange data with the input and output points
on the physical devices, or a communication interface that
allows the resources to communicate with external resources
in remote devices. The device model is illustrated in Figure
\ref{f:Device_Model}.

\begin{figure}
  \begin{center}
    \epsfig{file=images/functionblocks/Device_Model.eps,scale=.8} 
    \caption[Device Model]
            {Device Model{\protect ~\cite{iec:614991:2000}}}
    \label{f:Device_Model}
  \end{center}
\end{figure}

\subsection{Resource Model} 
A resource is an independent functional unit, contained in a
device which has independent control of its operation. The
resource can be created, configured, parameterized, started
up, deleted etc. without effecting other resources. Each
resource provides an interface to the communication systems
and tot the device specific process. The resource is
therefore, concerned with the mapping of data and event
flows which pass between function blocks in the local
resource to remove resource function blocks via the device
communication interfaces.

@@

\begin{figure}
  \begin{center}
    \epsfig{file=images/functionblocks/Resource_Model.eps,scale=.8} 
    \caption[Resource Model]
            {Resource Model{\protect ~\cite{iec:614991:2000}}}
    \label{f:Resource_Model}
  \end{center}
\end{figure}

\subsection{Application Model} 
An application is defined as a network of interconnected
function blocks, linked by data and event flows. An
application can be fragmented and distributed over many
resources. The resources ensure that the events are executed
in the correct serial order based on the causal
relationships specified by the applications.

@@

\begin{figure}
  \begin{center}
    \epsfig{file=images/functionblocks/Application_Model.eps, scale=.8}
    \caption[Application Model]
            {Application Model{\protect ~\cite{iec:614991:2000}}} 
    \label{f:Application_Model}
  \end{center}
\end{figure}

\subsection{Function Block Model} 
A function block is a functional unit of software that its
own data structure which can be manipulated by one or more
algorithms. The formal description of the data structure is
provided by the function block type definition.

@@

\begin{figure}
  \begin{center}
    \epsfig{file=images/functionblocks/Function_Block_Char.eps, scale=.8}
    \caption[Function Block Characteristics]
            {Function Block Characteristics{\protect ~\cite{iec:614991:2000}}}
    \label{f:Function_Block_Char}
  \end{center}
\end{figure}

\section{Function Block Types}
A function block type is defined by a type name, formal
definitions for the input events and data and the output
events and data. The type definition also includes the
internal behavior of the block. The notion of a function
block type is analogous to the class definition in
object-oriented programming while function block instances
correspond to the object instances of the classes.

\subsection{Basic Function Block Types} 
The behavior of basic function blocks is defined in terms of
algorithms that are invoked in response to input events. As
algorithms execute, they trigger output events to signal
their state to other blocks. The basic blocks hold their
state in terms of internal variables. Basic function blocks
also can communicate their state using input and output data
variables.

@@

\begin{figure}
  \begin{center}
    \epsfig{file=images/functionblocks/Basic_Function_Block.eps,scale=.8}
    \caption[Basic Function Block]
            {Basic Function Block{\protect ~\cite{iec:614991:2000}}} 
    \label{f:Basic_Block}
  \end{center}
\end{figure}

\subsection{Composite Function Blocks} 
The behavior of a composite function block is described in
terms of the event and data connections between its
component function blocks. A composite function block can be
used to hide the internal interactions between the component
blocks and presenting a simplified interface for use by the
outside world. A composite function block is illustrated in
Figure \ref{f:Composite_Block}.

\begin{figure}
  \begin{center}
    \epsfig{file=images/functionblocks/Composite_Function_Block.eps, scale=.8}
    \caption[Composite Function Block]
            {Composite Function Block{\protect ~\cite{iec:614991:2000}}} 
        \label{f:Composite_Block}
  \end{center}
\end{figure}

\subsection{Service Interface Function Blocks} 
Service interface function blocks provide an interface
between the function block domain and the external services.
Since a service interface function block is primarily
concerned with data exchange, it is defined using time
sequence diagrams. Examples of service interface function
blocks are communication blocks for publishing and
subscribing over a network and management blocks for
dynamically creating component blocks and event connections.

\subsection{Event Function Blocks} 
Event function blocks are some special basic function blocks
defined in the standard to ease the management of events.
These blocks are used for control, generation and detection
of events and can be used by other composite blocks. These
blocks are defined using an execution control chart or a
time sequence diagram. These blocks can be used for the
following purposes:
\begin{enumerate}
\item To split events to produce new events.
\item To allow events to be merged.
\item To permit the propagation of certain events.
\item To select between two or more events.
\item To delay an event by a given period.
\item To generate a stream of events at a regular interval.
\end{enumerate}

\section{Execution Model}
The execution of algorithms for basic function blocks is
invoked by the execution control portion of a function block
instance in response to events at event inputs. This
invocation takes the form of a request to the underlying
scheduling function of the associated resource to schedule
the execution of algorithm's operations. The model assumes
that the resource in which a function block exists provides
a scheduling function that ensures each phase of function
block execution occurs in the correct order and with the
correct priority.

@@

\begin{figure}
  \begin{center}
    \epsfig{file=images/functionblocks/Execution_Control_Chart.eps, scale=.8}
    \caption[Execution Model]
            {Execution Model{\protect ~\cite{iec:614991:2000}}}
    \label{f:Execution_Model}
  \end{center}
\end{figure}

@@
